### 2025-05-30 docs/article/Sections/benchmark.tex
Highlight Reactive versus Suspend for HtmlFlow templates and cross ref thos Listings in Results analysis.
review sections 4 and 5

\section{Benchmark Implementation}

The benchmark\footnote{\url{https://github.com/fmcarvalho/comparing-non-blocking-progressive-ssr}}
is designed with a modular architecture, separating the \textit{view} and
\textit{model} layers from the \textit{controller} layer ~\cite{Bucanek2009},
which allows for easy extension and integration of new template engines and
frameworks. It also includes a set of tests to ensure the correctness of
implementations and to validate the \textit{HTML} output. It includes two
different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class},
following the same model as
\textit{spring-webflux-comparing-template-engines}~\footnote{\url{https://github.com/xmlet/spring-webflux-comparing-template-engines}}.
The \texttt{Presentation} class represents a presentation with a title, speaker
name, and summary, while the \texttt{Stock} class represents a stock with a
name, URL, symbol, price, change, and ratio.

>>>>

\section{Benchmark Implementation}
\label{sec:bench}

This benchmark\footnote{\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}
is designed with a modular architecture, separating the \textit{view} and
\textit{model} layers from the \textit{controller} layer ~\cite{Bucanek2009},
which allows for easy extension and integration of new template engines and
frameworks. It also includes a set of tests to ensure the correctness of
implementations and to validate the \textit{HTML} output. It includes two
different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class}, 
following the proposal of the benchmarks~\cite{templatesbenchmark,springbenchmark}.
The \texttt{Presentation} class represents a presentation with a title, speaker
name, and summary, while the \texttt{Stock} class represents a stock with a
name, URL, symbol, price, change, and ratio.

### 2025-05-30 docs/article/Sections/results.tex
review sections 4 and 5

The results in Figure~\ref{fig:presentations-webflux-jmeter} show that when
using blocking template engines with a separate coroutine dispatcher, the
engines are unable to scale effectively beyond 4 concurrent users. In contrast,
the non-blocking engines scale efficiently up to 128 concurrent users, with
HtmlFlow achieving 4,487 requests per second. When blocking approaches are
executed in the context of Virtual Threads—thus enabling non-blocking I/O—the
engines also scale up to 128 concurrent users, with Jstachio using Virtual
Threads reaching 3,514 requests per second. The Thymeleaf implementation using
the reactive View Resolver driver also scales up to 128 concurrent users, but
achieves a lower throughput of approximately 2,559 requests per second.

@@

% The results for the Spring MVC implementation, shown in
% Figure~\ref{fig:presentations-springmvc-jmeter}, compare two approaches:
% \textit{Blocking}, which uses platform threads with \texttt{StreamingResponseBody},
% and \textit{Virtual}, which uses Virtual Threads. Both approaches scale
% effectively up to 128 concurrent users, with the Virtual Thread approach
% achieving a slightly higher throughput of 9,000 requests per second. However,
% these values are slightly lower than those observed in the Spring WebFlux
% implementation.
The results for the Spring MVC implementation, shown in
Figure~\ref{fig:presentations-springmvc-jmeter}, compare two synchronous
approaches: \textit{Blocking}, which uses platform threads with
\texttt{StreamingResponseBody}, and \textit{Virtual}, which uses Virtual
Threads. Since Spring MVC follows a thread-per-request architecture, the
asynchronous approaches—\textbf{Reactive} and \textbf{Suspendable}—described in
Section~\ref{sec:bench} are not applicable. Both the \textit{Blocking} and
\textit{Virtual} strategies scale effectively up to 128 concurrent users, with
the Virtual Threads approach achieving a slightly higher maximum throughput of
2,797 requests per second, compared to 2,498 requests per second for the
blocking approach. These results indicate that while Spring MVC can handle a
moderate level of concurrency, it does not reach the scalability of the
reactive or suspendable approaches available in Spring WebFlux. Furthermore,
Spring MVC does not enable Progressive Server-Side Rendering (PSSR) for the
tested templates, as previously discussed in Section~\ref{sec:bench}.

@@

The results for the Quarkus implementation, shown in
Figure~\ref{fig:presentations-quarkus-jmeter}, indicate that Quarkus handles
synchronous approaches more efficiently than Spring WebFlux. The blocking
engines scale up to 128 concurrent users, achieving up to 3744 requests per
second. When using Virtual Threads, the throughput increases even further,
reaching 4856 requests per second. This demonstrates that Quarkus's
implementation of Virtual Threads is effective for enabling PSSR\@, and
comparable to the \textit{Suspendable} and \textit{Reactive} used approaches in
Spring WebFlux in terms of scalability and throughput.

Additionally, \textit{HtmlFlow-Rx}, a reactive implementation of the HtmlFlow
template engine (equivalent to the approach shown in
Listing~\ref{lst:presentation-observable}) that utilizes Quarkus's reactive
programming model, achieved a lower throughput than the the \textit{Blocking}
and \textit{Virtual} approaches—3088 requests per second. This demonstrates
that Quarkus's reactive programming model is effective for enabling PSSR\@,
although it does not achieve the same level of performance as the same approach
in Spring WebFlux.

@@

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but replace the data model
with the more complex Stock class, including 20 instances. When using this data
model, we observe that the scalability of the engines is not significantly
affected, however the throughput is reduced across all engines. When compared
to the Presentation benchmark, we observe that particularly the approaches
using Virtual Threads took a greater performance hit when compared to the
\textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
Virtual Threads now achieving 1509 requests per second, compared to the 1,750
requests per achieved by the Thymeleaf implementation using the reactive View
Resolver driver.

@@

The results shown in Figure~\ref{fig:stocks-springmvc-jmeter} indicate that the
Spring MVC implementation using the blocking approach with
\texttt{StreamingResponseBody} achieves a throughput of up to 1,916 requests
per second, with no significant improvement observed when using Virtual
Threads. Both approaches scale effectively up to 128 concurrent users. Although
these approaches achieve higher throughput in Spring MVC than in Spring
WebFlux, their overall performance remains lower than that of the reactive and
suspendable approaches.

@@

The results depicted in Figure~\ref{fig:stocks-quarkus-jmeter} show that the
Quarkus implementation scales effectively up to 128 concurrent users, achieving
performance comparable to the Spring WebFlux implementation. The blocking
approach reaches a throughput of 3,019 requests per second, while the Virtual
Threads approach achieves a throughput of 3,357 requests per second.

In addition to the blocking engines, the \textit{HtmlFlow-Rx} approach also
achieves a throughput of 1760 requests per second, indicating that this
approach achieves lower performance in Quarkus than in Spring WebFlux, where it
reached 3,026 requests per second, as previously mentioned.

The results of the benchmarks show that non-blocking engines, through the use
of reactive programming, Kotlin coroutines, or Java virtual threads, are able
to scale effectively up to 128 concurrent users. Out of all the tested
frameworks, Spring Webflux showed itself the most effective at enabling PSSR,
mostly due to its native support for Publish and Subscriber interfaces, which
allow for HTML content to be progressively streamed to the client. Quarkus also
enabled PSSR effectively, but it required additional configuration of the
\texttt{OutputBuffer} size to achieve the same results as Spring Webflux. The
Spring MVC implementation, on the other hand, did not enable PSSR for the
tested templates.

Additionally, the results show that approaches using Virtual Threads are able
to scale as effectively as those using reactive programming or Kotlin
coroutines, allowing \textit{external} DSLs to be used in a non-blocking
context, effectively enabling PSSR\@.

>>>>

The results in Figure~\ref{fig:presentations-webflux-jmeter} show that when
using blocking template engines with a separate coroutine dispatcher, the
engines are unable to scale effectively beyond 4 concurrent users. In contrast,
\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4,487
requests per second. When blocking approaches are executed in the context of
Virtual Threads—thus enabling non-blocking I/O—the engines scale up to 64
concurrent users, with Jstachio using Virtual Threads reaching 3,514 requests
per second. The Thymeleaf implementation using the reactive View Resolver
driver scales up to 32 concurrent users, achieving a lower maximum throughput
of 2,559 requests per second.

@@

% The results for the Spring MVC implementation, shown in
% Figure~\ref{fig:presentations-springmvc-jmeter}, compare two approaches:
% \textit{Blocking}, which uses platform threads with \texttt{StreamingResponseBody},
% and \textit{Virtual}, which uses Virtual Threads. Both approaches scale
% effectively up to 128 concurrent users, with the Virtual Thread approach
% achieving a slightly higher throughput of 9,000 requests per second. However,
% these values are slightly lower than those observed in the Spring WebFlux
% implementation.
The results for the Spring MVC implementation, shown in
Figure~\ref{fig:presentations-springmvc-jmeter}, compare two synchronous
approaches: \textit{Blocking}, which uses platform threads with
\texttt{StreamingResponseBody}, and \textit{Virtual}, which uses Virtual
Threads. Since Spring MVC follows a thread-per-request architecture, the
asynchronous approaches—\textbf{Reactive} and \textbf{Suspendable}—described in
Section~\ref{sec:bench} are not applicable. Both the \textit{Blocking} and
\textit{Virtual} strategies scale effectively up to 32 concurrent users, with
the Virtual Threads approach achieving a slightly higher maximum throughput of
2,797 requests per second, compared to 2,498 requests per second for the
blocking approach. These results indicate that while Spring MVC can handle a
moderate level of concurrency, it does not reach the scalability of the
reactive or suspendable approaches available in Spring WebFlux. Furthermore,
Spring MVC does not enable Progressive Server-Side Rendering (PSSR) for the
tested templates, as previously discussed in Section~\ref{sec:bench}.

@@

The results for the Quarkus implementation, shown in
Figure~\ref{fig:presentations-quarkus-jmeter}, indicate that Quarkus handles
synchronous approaches more efficiently than Spring WebFlux. The blocking
engines scale up to 64 concurrent users, achieving up to 3,744 requests per
second. When using Virtual Threads, the throughput increases even further,
reaching 4,856 requests per second, allowing scalability up to 128 users. This
demonstrates that Quarkus's implementation of Virtual Threads is effective for
enabling PSSR\@, and comparable to the \textit{Suspendable} and
\textit{Reactive} approaches used in Spring WebFlux in terms of scalability and
throughput.

Additionally, \textit{HtmlFlow-Rx}, a reactive implementation of the HtmlFlow
template engine (equivalent to the approach shown in
Listing~\ref{lst:presentation-observable}) that utilizes Quarkus's reactive
programming model, achieved a lower throughput than the the \textit{Blocking}
and \textit{Virtual} approaches—3,088 requests per second. This demonstrates
that Quarkus's reactive programming model is effective for enabling PSSR\@,
although it does not achieve the same level of performance or scalability as
the same approach in Spring WebFlux, stagnating after 32 concurrent users.

@@

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but with a more complex data
model: the Stock class, which includes 20 instances and approximately two
times as many data bindings. With this data model, the scalability of the
engines remains largely unchanged; however, throughput is reduced across all
engines. Compared to the Presentation benchmark, the approaches using Virtual
Threads experience a more pronounced decrease in performance relative to the
\textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
Virtual Threads now achieving 1,509 requests per second, compared to 1,750
requests per second achieved by the Thymeleaf implementation using the reactive
View Resolver driver.

@@

The results shown in Figure~\ref{fig:stocks-springmvc-jmeter} indicate that the
Spring MVC implementation using the blocking approach with
\texttt{StreamingResponseBody} achieves a throughput of up to 1,916 requests
per second, with no significant improvement observed when using Virtual
Threads. Both approaches scale effectively up to 64 concurrent users. Although
these approaches achieve higher throughput in Spring MVC than in Spring
WebFlux, their overall performance remains lower than that of the reactive and
suspendable approaches.

@@

The results depicted in Figure~\ref{fig:stocks-quarkus-jmeter} show that the
Quarkus synchronous approaches scale effectively up to 128 concurrent users,
achieving performance comparable to the Spring WebFlux implementation. The
blocking approach reaches a throughput of 3,019 requests per second, while the
Virtual Threads approach achieves a throughput of 3,357 requests per second. In
addition to the synchronous engines, the \textit{HtmlFlow-Rx} approach also
achieves a throughput of 1760 requests per second, indicating that this
approach achieves lower performance in Quarkus than in Spring WebFlux, where it
reached 3,026 requests per second, as previously mentioned.

The results of the benchmarks show that non-blocking engines—whether using
reactive programming, Kotlin coroutines, or Java Virtual Threads—are able to
scale effectively, supporting between 32 and 128 concurrent users depending on
the approach and framework. Out of all the tested frameworks, Spring Webflux
showed itself the most effective at enabling PSSR, mostly due to its native
support for Publish and Subscriber interfaces, which allow for HTML content to
be progressively streamed to the client. Quarkus also enabled PSSR effectively,
but it required additional configuration of the \texttt{OutputBuffer} size to
achieve the same results as Spring Webflux. The Spring MVC implementation, on
the other hand, did not enable PSSR for the tested templates.

% Additionally, the results show that approaches using Virtual Threads are able
% to scale as effectively as those using reactive programming or Kotlin
% coroutines, allowing \textit{external} DSLs to be used in a non-blocking
% context, effectively enabling PSSR\@.
